---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# Malawi Groundwater Monitoring Time-Series Dataset (2024–2025)

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

[![DOI](https://zenodo.org/badge/1167601334.svg)](https://doi.org/10.5281/zenodo.18798409)

<!-- badges: end -->

This dataset contains groundwater monitoring data collected from
monitoring wells across 10 districts in Malawi between 2024 and 2025.
The data were captured using automated data loggers installed in
monitoring wells and were collected and managed by BASEflow.

The dataset provides time-series measurements of key groundwater
parameters, enabling detailed analysis of aquifer behavior across
multiple geographic locations.

Variables Included

-   Date – Date of measurement

-   Waterpoint Name – Name of the monitoring site

-   District – Administrative district where the monitoring well is
    located

-   Latitude & Longitude – Geographic coordinates of the monitoring well

-   Source – Data collection method (automated data logger)

-   Water Level – Groundwater level measurement (typically meters below
    ground level, depending on installation reference)

-   Temperature – Groundwater temperature (°C)

-   Conductivity – Electrical conductivity (µS/cm), indicating dissolved
    ion concentration and groundwater quality characteristics

The use of automated loggers ensures high-frequency, consistent, and
reliable measurements suitable for time-series analysis and
hydrogeological assessment.

1.  **Purpose and Use Cases 1. Groundwater Resource Monitoring**

-   Tracking spatial and temporal groundwater level variations across
    districts

-   Assessing seasonal recharge and depletion patterns

-   Identifying long-term aquifer trends

2.  **Water Quality Surveillance**

-   Monitoring conductivity trends as a proxy for salinity and
    mineralization

-   Detecting potential contamination or quality shifts

3.  **Climate and Drought Analysis**

-   Supporting drought early warning systems

-   Evaluating groundwater resilience to climate variability

4.  **Infrastructure Management**

-   Informing borehole design and pump installation depths

-   Supporting preventive maintenance planning

-   Assessing borehole performance over time

5.  **Hydrogeological Research and Modelling**

-   Input data for groundwater flow and recharge models

-   Calibration of aquifer simulations

-   Comparative inter-district hydrogeological analysis

6.  **Policy, Regulation, and Planning**

-   Evidence base for district-level and national water resource
    planning

-   Supporting groundwater abstraction regulation

-   Informing investment decisions in rural and urban water supply

**Potential Users**

-   Ministry responsible for Water and district water offices

-   Hydrologists and hydrogeologists

-   WASH sector NGOs and implementing partners

-   Academic and research institutions

-   Climate and environmental analysts

-   Development partners supporting water security and resilience
    programs

This dataset provides a structured, multi-district groundwater evidence
base to support sustainable groundwater management and water security
planning in Malawi.

## Installation

You can install the development version of mwgroundwaterdata from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/mwgroundwaterdata")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/mwgroundwaterdata/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to This dataset contains groundwater
monitoring data collected from monitoring wells across 10 districts in
Malawi between 2024 and 2025. The data were captured using automated
data loggers installed in monitoring wells and were collected and
managed by BASEflow.

```{r}
library(mwgroundwaterdata)
```

### metadata

The dataset `mwgroundwaterdata` contains `r nrow(mwgroundwaterdata)`
observations and `r ncol(mwgroundwaterdata)` variables

```{r}
mwgroundwaterdata |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "mwgroundwaterdata.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

```{r}
library(mwgroundwaterdata)

# Visualization: Geospatial Map
# Import the libraries to be used
library(tidyverse)
library(lubridate)
library(leaflet)

# Create summary dataset INSIDE the README
well_summary <- mwgroundwaterdata %>%
  group_by(waterpoint_name, latitude, longitude, district) %>%
  summarise(
    avg_water_level = mean(water_level, na.rm = TRUE),
    avg_conductivity = mean(conductivity, na.rm = TRUE),
    .groups = "drop"
  )

leaflet(well_summary) %>%
  addTiles() %>%  # OpenStreetMap tiles
  addCircleMarkers(~longitude, ~latitude,
                   radius = ~avg_conductivity/400,
                   color = ~colorNumeric("plasma", avg_water_level)(avg_water_level),
                   popup = ~paste0(waterpoint_name, "<br>Avg Water Level: ", round(avg_water_level,2),
                                   "<br>Avg Conductivity: ", round(avg_conductivity,1))) %>%
  addLegend("bottomright",
            pal = colorNumeric("plasma", well_summary$avg_water_level),
            values = well_summary$avg_water_level,
            title = "Avg Water Level (m)")

```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/%7B%7B%7Bpackagename%7D%7D%7D/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("mwgroundwaterdata")
```
